---
title: "Assessment of fisheries impact in the Dutch part of the Northsea, 2009. The BFIAT approach"
subtitle: "part 3 - gridding fishing impact"
author: "karline soetaert"
date: 13-03-2026
toc: true
toc-depth: 3
toc-expand: true
code-fold: show
format: 
  html:
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
abstract: "Fishing impact on macrofaunal biomass was assessed based on monitoring data of the Dutch part of the Northsea, from 2009. Here we use these data to generate full coverage maps. We do this by simple kriging, and then masking the land. After that, netCDF files are generated."    
---

For mapping we use three R packages: raster (Hijmans, 2025), gstat (Pebesma 2004), and sp (Pebesma and Bivan, 2005): 

```{r, warning = FALSE, message=FALSE}
require(raster)
require(sp)
require(gstat)
```

```{r, warning = FALSE, message=FALSE}
require(RColorBrewer)
col.fau <- colorRampPalette(c( "darkblue", "#3E9188",  "#FEEEA0"))(100)
```

The fishing impact for 100 stations in the Dutch part of the Northsea for the year 2009 were modeled in previous tutorial, and its output written to data.frame called *station_bfiat*.

The data.frame contains:

* the *station* names and positions (*x*, *y*), 
* two fishing parameters: the annual swept area ratio (*sar*, $yr^{-1}$), and the gear penetration depth (*gpd*, $cm$),
* the observed macrofaunal *biomass* ($gAFDW~m^{-2}$) and *density* ($ind~m^{-2}$) for the year 2009, 
* the carrying capacity in terms of biomass and density (*K_biomass*, *K_density*); this is the biomass or density that would be present in the absence of fishing,
* the fishing impact on biomass, estimated as:
      - the current biomass as a fraction of the carrying capacity(*DK_biomass = biomass/K_biomass*),
      - the fractional loss of biomass *fraction_biomass_lost = 1 = DK_biomass*,
* the current bioturbation potential (*BPc*), and the bioturbation potential at carrying capacity (*K_BPc*),
* the fishing impact on bioturbation potential: 
      - the current bioturbation potential as a fraction of the unfished potential (*DK_BPc = BPc/K_BPc*), 
      - the fractional loss of bioturbation potential (*fraction_BPc_lost = 1 - DK_BPc*), 
* the current bioirrigation potential (*IPc*), and the bioirrigation potential at carrying capacity (*K_IPc*),
* the fishing impact on bioirrigation potential: 
      - the current bioirrigation potential as a fraction of the unfished potential  (*DK_IPc = IPc/K_IPc*),
      - the fractional loss of bioirrigation potential (*fraction_IPc_lost = 1 - DK_IPc*). 

```{r}
load("../data/derived_data/station_bfiat.rda")
head(station_bfiat)
```

# Kriging

For ease of interpolation, a function is created:

```{r}
interp <- function(df = station_bfiat,                  # data.frame with data 
                   ext = extent(2.6, 6.65, 51.6, 55.5), # extension of the grid
                   nc       = 350, 
                   nc.grid  = 400,        # number of columns in grid
                   z,                     # variable name from df
                   mod      = vgm("Mat"), # model to use
                   log      = FALSE,      # logarithm transform
                   plot.vgm = TRUE,       # plot the variogram
                   width    = NA, 
                   ...
                   ){
  cat(z, "\n")

# number of rows used
nr  <- round(abs(ext[4] - ext[3]) / (abs(ext[2] - ext[1]) / nc), 0)

# clean up the data
# ---------------------  
  df   <- df[df$x >= ext[1] & df$x <= ext[2]  &
             df$y >= ext[3] & df$y <= ext[4], ]

  coordinates(df) <- ~x+y
  crs(df)         <- "+proj=longlat +datum=WGS84"
  
  if (nrow(zerodist(df)) > 0)
    df      <- df[-zerodist(df)[,1],]
  
# create kriging grid
# ---------------------  
  rg.x    <- abs(ext[1] - ext[2])
  rg.y    <- abs(ext[3] - ext[4])

  nr.grid <- ceiling(rg.y/(rg.x/nc.grid))
  grd     <- GridTopology(
      cellcentre.offset = c(ext[1],       ext[3]),        
      cellsize          = c(rg.x/nc.grid, rg.x/nc.grid), 
      cells.dim         = c(nc.grid,      nr.grid)       
  )
  
  grd      <- SpatialGrid(grd)
  crs(grd) <- "+proj=longlat +datum=WGS84"

# model formula
# ---------------------  
  if (log == FALSE){
    f <- as.formula(paste(z, "~1", sep = ""))
  
  } else {
    f <- as.formula(paste("log(", z, "+1)", "~1", sep = ""))
  }

# calculate variogram
# ---------------------  
  
  if (is.na(width)) 
    vgm  <- variogram(f, data = df, ...)
  else
    vgm  <- variogram(f, data = df, width = width, ...)
  
  fit  <- fit.variogram(vgm, model = mod)
  print(fit)
  if (plot.vgm == TRUE){
    plot(vgm$dist, vgm$gamma, 
         ylim = c(0, max(vgm$gamma)),
         xlab = "Distance", ylab = "Semivariance")
    lines(variogramLine(fit, maxdist = max(vgm$dist)))
  }
  
# interpolate
# ---------------------  
  k <- krige(f, df, grd, model = fit)
  k <- as(k, "SpatialPointsDataFrame")
  r <- rasterize(k, 
                 raster(xmn = ext[1], xmx = ext[2],
                        ymn = ext[3], ymx = ext[4], 
                        nrows = nr, ncols = nc), 
                 fun = mean)
  names(r)[c(2, 3)] <- c("pred", "var")
  r$pred[r$pred < 0] <- 0

  return(r)

}
```

```{r}

# variable names in input to krige
nn       <- c("sar",  
              "fraction_biomass_lost", 
              "fraction_BPc_lost", 
              "fraction_IPc_lost")

longname <- c("swept area ratio", 
              "fraction biomass lost by fishing", 
              "fraction bioturbation lost by fishing", 
              "fraction bioirrigation lost by fishing")
units    <- c("/yr", "-", "-", "-")
```

```{r, fig.width=10, fig.height=10}
par(mfrow = c(2, 2))

ext = extent(2.6, 6.65, 51.6, 55.5)

# number of columns used
nc  <- 350

# number of rows used
nr  <- round(abs(ext[4] - ext[3]) / (abs(ext[2] - ext[1]) / nc), 0)

# create bricks, 4 layers
b      <- brick(nl = 4, ext, nrows = nr, ncols = nc)

b[[1]] <- interp(z   = "sar", 
                 mod = vgm(
                      model  = "Exc",
                      nugget = 2,
                      psill  = 6,
                      range  = 50)
            )$pred
b[[2]] <- interp(z   = "fraction_biomass_lost", 
                 mod = vgm(
                      model  = "Lin",
                      nugget = 0.01,
                      psill  = 0.025,
                      range  = 100)
            )$pred

b[[3]] <- interp(z   = "fraction_BPc_lost", 
                 mod = vgm(
                      model  = "Lin",
                      nugget = 0.01,
                      psill  = 0.05,
                      range  = 100)
            )$pred

b[[4]] <- interp(z   = "fraction_IPc_lost", 
                 mod = vgm(
                      model  = "Lin",
                      nugget = 0.005,
                      psill  = 0.01,
                      range  = 100), 
                 width = 10
            )$pred
names(b) <- longname
```

# Masking and plotting

A number of shapefiles have been downloaded and put in a dataset.
They are:

* eez_nl, the border of the Dutch exclusive economic zone in the Northsea
* cst_nl, the coast around the Dutch coast
* ws_nl, the wadden sea area

These shape files are used to mask the gridded data outside of the dutch eez area.

Masking is done in a number of steps

```{r}
load(file = "../data/raw_data/shapefiles.rda") # eez_nl, cstl_nl, ws_nl

mask_to_mwtl <- function(b){
  BB <- mask(b,  eez_nl)                     # dutch economic zone
  BB <- mask(BB, cstl_nl,   inverse = TRUE)  # only offshore of coast 
  BB <- mask(BB, ws_nl,     inverse = TRUE)  # remove Waddenzee
  crs(BB) <-  "EPSG:4326" 
  BB
}

plot_brick <- function(b, col = col.fau, legend = "-", ...){
  plot (b, col = col, legend.args = list(text = legend), ...)
  plot (cstl_nl,   col = "grey", add = TRUE)
  lines(cstl_nl,   col = "black")
  lines(eez_nl,    col = "black")
  box()
}

for (i in 1:4) 
  b[[i]] <- mask_to_mwtl(b[[i]])


```


```{r, fig.width=10, fig.height=10}
require(shape)
par(mfrow = c(2,2), oma = c(0,0,0,2), las = 1)

plot_brick(b[[1]],  main = "swept area ratio", 
           legend = '/yr')
  writelabel(LETTERS[1])

plot_brick(b[[2]],  main = "fraction biomass lost by fishing", legend = '-')
  writelabel(LETTERS[2])
plot_brick(b[[3]],  main = "fraction bioturbation lost by fishing", legend = '-')
  writelabel(LETTERS[3])
plot_brick(b[[4]],  main = "fraction bioirrigation lost by fishing", legend = '-')
  writelabel(LETTERS[4])

```

# Save as separate netcdf files

```{r}

for (i in 1:4)
writeRaster(x         = b[[i]], 
            filename  = paste("../data/derived_data/", 
                              nn[i], 
                              ".nc", sep = ""), 
            overwrite = TRUE, 
            format    = "CDF",  
            varname   = nn[i], 
            varunit   = units[i], 
            longname  = longname[i], 
            xname     = "longitude", 
            yname     = "latitude")
```


In next document, one NETCDF file will be generated.

# References

Hijmans R (2025). _raster: Geographic Data Analysis and Modeling_. doi:10.32614/CRAN.package.raster
  <https://doi.org/10.32614/CRAN.package.raster>, R package version 3.6-32,
  <https://CRAN.R-project.org/package=raster>.
  
Pebesma, E.J., 2004. Multivariable geostatistics in S: the gstat package. Computers & Geosciences, 30:
  683-691.

Pebesma E, Bivand R (2005). “Classes and methods for spatial data in R.” _R News_, *5*(2), 9-13.
  <https://CRAN.R-project.org/doc/Rnews/>.
  
