---
title: "Assessment of fisheries impact in the Dutch part of the Northsea, 2009. The BFIAT approach"
subtitle: "part 4 - creating the NETCFD file"
author: "karline soetaert"
date: 13-03-2026
toc: true
toc-depth: 3
toc-expand: true
code-fold: show
format: 
  html:
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
abstract: "Fishing impact on macrofaunal biomass was assessed based on monitoring data of the Dutch part of the Northsea, from 2009. We use the kriged data to generate one fully documented netCDF file"    
---

# Opening separate NETCDF files and reading their contents

```{r}
library(RNetCDF)
ncipc <- open.nc("../data/derived_data/fraction_IPc_lost.nc")
ncbpc <- open.nc("../data/derived_data/fraction_BPc_lost.nc")
ncbio <- open.nc("../data/derived_data/fraction_biomass_lost.nc")
nc    <- create.nc("../product/fishing_impact_2009_bfiat.nc")

ipc <- var.get.nc(ncipc, "fraction_IPc_lost")
bpc <- var.get.nc(ncbpc, "fraction_BPc_lost")
bio <- var.get.nc(ncbio, "fraction_biomass_lost")
lon <- var.get.nc(ncipc, "longitude")
lat <- var.get.nc(ncipc, "latitude")
```

# Defining the dimensions

```{r}
# ----------------
# longitude
# ----------------
dim.def.nc(nc, dimname   = "lon", 
               dimlength = length(lon)) 

var.def.nc(nc, varname    = "lon", 
               vartype    = "NC_DOUBLE", 
               dimensions = "lon")

att.put.nc(nc, variable = "lon", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "degrees_east")

att.put.nc(nc, variable = "lon", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "longitude")

att.put.nc(nc, variable = "lon", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value    = "Longitude")

var.put.nc(nc, variable = "lon", 
               data     = lon) 

# Check
#var.get.nc(nc, variable = "lon")

# ----------------
# latitude
# ----------------
# Define lat dimension
dim.def.nc(nc, dimname   = "lat", 
               dimlength = length(lat)) 

# Define lat variable
var.def.nc(nc, varname    = "lat", 
               vartype    = "NC_DOUBLE", 
               dimensions = "lat")

# Add attributes
att.put.nc(nc, variable = "lat", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "degrees_north")

att.put.nc(nc, variable = "lat", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "latitude")

att.put.nc(nc, variable = "lat", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value    = "Latitude")

# Put data
var.put.nc(nc, variable = "lat", 
               data     = lat) 

# Check
#var.get.nc(nc, variable = "lat")

# ----------------
# time
# ----------------
# Define time dimension
dim.def.nc(nc, dimname = "time", dimlength = 1) 

# Define time variable
var.def.nc(nc, varname    = "time", 
               vartype    = "NC_DOUBLE", 
               dimensions = "time")

# Add attributes
att.put.nc(nc, variable = "time", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "time")

att.put.nc(nc, variable = "time", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value    = "Time")

att.put.nc(nc, variable = "time", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "year")

att.put.nc(nc, variable = "time", 
               name     = "calendar", 
               type     = "NC_CHAR", 
               value    = "gregorian")

# Put data
var.put.nc(nc, variable = "time", 
               data     = 2009.)

# Check
#var.get.nc(nc, variable = "time")

```

# variables

## fraction biomass lost due to bottom trawling

```{r}
var.def.nc(nc, varname   = "fraction_biomass_lost", 
               vartype   = "NC_DOUBLE", 
               dimension = c("lon", "lat"))

att.put.nc(nc, variable = "fraction_biomass_lost", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "fraction_biomass_lost_to_fishing")

att.put.nc(nc, variable = "fraction_biomass_lost", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value = "fraction of macrofaunal biomass lost due to bottom trawling for 2009, estimated as 1-D/K (BFIAT approach)")

att.put.nc(nc, variable = "fraction_biomass_lost", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "-")

att.put.nc(nc, variable = "fraction_biomass_lost", 
               name     = "formula", 
               type     = "NC_CHAR", 
               value    = "(1 - Biomass_observed/Biomass_carrying_capacity)")

att.put.nc(nc, variable = "fraction_biomass_lost", 
               name     = "_FillValue", 
               type     = "NC_DOUBLE", 
               value    = -99999.0)

# Put data
var.put.nc(nc, variable = "fraction_biomass_lost", 
               data     = bio)

# Check
#var.get.nc(nc, variable = "fraction_biomass_lost")
```

## fraction bioturbation potential lost due to bottom trawling

```{r}
var.def.nc(nc, varname   = "fraction_BPc_lost", 
               vartype   = "NC_DOUBLE", 
               dimension = c("lon", "lat"))

att.put.nc(nc, variable = "fraction_BPc_lost", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "fraction_bioturbation_lost_to_fishing")

att.put.nc(nc, variable = "fraction_BPc_lost", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value    = "fraction of bioturbation potential (BPc) lost due to bottom trawling for 2009, estimated as 1-BPc/K_BPc (BFIAT approach)")

att.put.nc(nc, variable = "fraction_BPc_lost", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "-")


att.put.nc(nc, variable = "fraction_BPc_lost", 
               name     = "formula", 
               type     = "NC_CHAR", 
               value    = "(1 - BPc_observed/BPc_carrying_capacity)")

att.put.nc(nc, variable = "fraction_BPc_lost", 
               name     = "_FillValue", 
               type     = "NC_DOUBLE", 
               value    = -99999.0)

# Put data
var.put.nc(nc, variable = "fraction_BPc_lost", 
               data     = bpc)

# Check
#var.get.nc(nc, variable = "fraction_biomass_lost")
```

## fraction bioirrigation potential lost due to bottom trawling


```{r}
var.def.nc(nc, varname   = "fraction_IPc_lost", 
               vartype   = "NC_DOUBLE", 
               dimension = c("lon", "lat"))

att.put.nc(nc, variable = "fraction_IPc_lost", 
               name     = "standard_name", 
               type     = "NC_CHAR", 
               value    = "fraction_bioirrigation_lost_to_fishing")

att.put.nc(nc, variable = "fraction_IPc_lost", 
               name     = "long_name", 
               type     = "NC_CHAR", 
               value    = "fraction of bioirrigation potential (IPc) lost due to bottom trawling for 2009, estimated as 1-IPc/K_IPc (BFIAT approach)")

att.put.nc(nc, variable = "fraction_IPc_lost", 
               name     = "units", 
               type     = "NC_CHAR", 
               value    = "-")


att.put.nc(nc, variable = "fraction_IPc_lost", 
               name     = "formula", 
               type     = "NC_CHAR", 
               value    = "(1 - IPc_observed/IPc_carrying_capacity)")

att.put.nc(nc, variable = "fraction_IPc_lost", 
               name     = "_FillValue", 
               type     = "NC_DOUBLE", 
               value    = -99999.0)

# Put data
var.put.nc(nc, variable = "fraction_IPc_lost", 
               data     = ipc)

# Check
#var.get.nc(nc, variable = "fraction_biomass_lost")
```

## Non-dimensional crs variable

```{r}
# Define non-dimensional crs variable 
var.def.nc(nc, varname = "crs", vartype = "NC_CHAR", dimensions = NA)

# Add attributes
att.put.nc(nc, variable = "crs", 
               name     = "long_name", type = "NC_CHAR", 
               value    = "Coordinate Reference System")
att.put.nc(nc, variable = "crs", name = "geographic_crs_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "grid_mapping_name", type = "NC_CHAR", value = "latitude_longitude")
att.put.nc(nc, variable = "crs", name = "reference_ellipsoid_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "horizontal_datum_name", type = "NC_CHAR", value = "WGS 84")
att.put.nc(nc, variable = "crs", name = "prime_meridian_name", type = "NC_CHAR", value = "Greenwich")
att.put.nc(nc, variable = "crs", name = "longitude_of_prime_meridian", type = "NC_DOUBLE", value = 0.)
att.put.nc(nc, variable = "crs", name = "semi_major_axis", type = "NC_DOUBLE", value = 6378137.)
att.put.nc(nc, variable = "crs", name = "semi_minor_axis", type = "NC_DOUBLE", value = 6356752.314245179)
att.put.nc(nc, variable = "crs", 
               name     = "inverse_flattening", 
               type = "NC_DOUBLE", value = 298.257223563)
att.put.nc(nc, variable = "crs", 
               name     = "spatial_ref", type = "NC_CHAR", 
               value    = 'GEOGCS[\"WGS 84\",DATUM[\"WGS_1984\",SPHEROID[\"WGS 84\",6378137,298.257223563]],PRIMEM[\"Greenwich\",0],UNIT[\"degree\",0.0174532925199433,AUTHORITY[\"EPSG\",\"9122\"]],AXIS[\"Latitude\",NORTH],AXIS[\"Longitude\",EAST],AUTHORITY[\"EPSG\",\"4326\"]]')

att.put.nc(nc, variable = "crs", 
               name     = "GeoTransform", type = "NC_CHAR", 
               value    = '-180 0.08333333333333333 0 90 0 -0.08333333333333333 ')
```

# attributes


```{r}
attributes <- list(
  title = "Trawling impact on macrofauna in the Dutch part of the Northsea, for 2009",
  summary = "This estimates the impact of trawling on macrofaunal communities in the Dutch part of the North sea, in terms of total biomass, bioturbation potential and bioirrigation potential. Impact on biomass is represented as one minus the ratio of biomass observed in 2009 over the carrying capacity: (1 - B/K), where B is the observed biomass from 2009 and K the estimated biomass in the absence of fishing. The Dutch monitoring data (MWTL) for 2009 were extracted from EMODnet biology; fishing intensity was extracted from ICES. Data creation used the R-packages Btrait and Bfiat; the procedure is described in Beauchard and Soetaert (2026).",                       
  source = "Beauchard, Olivier and Soetaert, Karline, 2026 Bottom fishing assessment tool: An R package to model the effects of bottom trawling on the marine benthos. Ecological Applications(36) 5, e70282. URL https://doi.org/10.1002/eap.70282",
  license = "CC-BY",
  id = "",
  history = "https://github.com/EMODnet/Bfiat_application_MWTL_2009",
  source = "https://github.com/EMODnet/Bfiat_application_MWTL_2009",

  keywords = c("fishing", "impact", "macrobenthos"),
  date_created = as.character(Sys.Date()),
  creator_name = "Karline soetaert",
  creator_email = "karline.soetaert@nioz.nl",
  creator_url = "www.nioz.nl",
  institution = "Netherlands Institute for Sea Research (NIOZ)",
  project = "EMODnet-Biology",
  publisher_name = "EMODnet-Biology",                 
  publisher_email = "bio@emodnet.eu",                
  publisher_url = "https://emodnet.ec.europa.eu/en/biology",  
  geospatial_bounds = "",              
  geospatial_bounds_crs = "",          
  geospatial_bounds_vertical_crs = "", 

  geospatial_lat_min = min(lat),
  geospatial_lat_max = max(lat),
  geospatial_lon_min = min(lon),
  geospatial_lon_max = max(lon),
  time_coverage_start = "2009",            
  time_coverage_end = "2009",              
  time_coverage_duration = 1,         
  time_coverage_resolution = "year",   
  sea_name = "North Sea",                       
  creator_institution = "Netherlands Institute for Sea Research (NIOZ)",           
  geospatial_lat_units = "degrees_north",           
  geospatial_lon_units = "degrees_east",           
  comment = "Uses R-packages Bfiat and Btrait",
  license = "CC-BY", 
  citation = "Beauchard, Olivier and Soetaert, Karline, 2026 Bottom fishing assessment tool: An R package to model the effects of bottom trawling on the marine benthos. Ecological Applications(36) 5, e70282. URL https://doi.org/10.1002/eap.70282",
  acknowledgement = "European Marine Observation Data Network (EMODnet) Biology project CINEA/EMFAF/2022/3.5.2/SI2.895681, funded by the European Union under Regulation (EU) No 508/2014 of the European Parliament and of the Council of 15 May 2014 on the European Maritime and Fisheries Fund"
)

add_global_attributes <- function(nc, attributes){
  
  stopifnot(is.list(attributes))
  
  for(i in 1:length(attributes)){
    if(is.character(attributes[[i]])){
      type <- "NC_CHAR"
    }else if(is.numeric(attributes[[i]])){
      type <- "NC_DOUBLE"
    }
    att.put.nc(nc, variable = "NC_GLOBAL", name = names(attributes[i]), type = type, value = attributes[[i]])
  }
  sync.nc(nc)
}

# Add attributes
add_global_attributes(nc, attributes)
add_global_attributes <- function(nc, attributes){
  
  stopifnot(is.list(attributes))
  
  for(i in 1:length(attributes)){
    if(is.character(attributes[[i]])){
      type <- "NC_CHAR"
    }else if(is.numeric(attributes[[i]])){
      type <- "NC_DOUBLE"
    }
    att.put.nc(nc, variable = "NC_GLOBAL", name = names(attributes[i]), type = type, value = attributes[[i]])
  }
  sync.nc(nc)
}

# Add attributes
add_global_attributes(nc, attributes)
sync.nc(nc)
print.nc(nc)
close.nc(nc)
```

# References

 Michna P, Woods M (2026). _RNetCDF: Interface to 'NetCDF' Datasets_.
  doi:10.32614/CRAN.package.RNetCDF <https://doi.org/10.32614/CRAN.package.RNetCDF>, R
  package version 2.11-2, <https://CRAN.R-project.org/package=RNetCDF>.
  